---
title: "TERRG_H2_sample_estimate"
author: "Sean Kinard"
date: "2/17/2022"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r, include=FALSE, echo=FALSE}
# load packages
library(tidyverse)
library('sjmisc')
library(lubridate)
library(ggpubr)
library(rstatix)

# set work directory
setwd('C:\\Users\\s2kin\\Dropbox\\Research\\Manuscipts\\Diss.5_Isotope\\RAPID-TERRG-Combined\\')

# load data
d <- read_csv('data/TERRG_CNS_combined.csv') %>% 
  filter(site %in% c("AR","EM", "GC", "MR", "PD", "PL", "SF", "TR", "WM") )

d$site <- factor(d$site, levels = c('TR', 'SF', 'AR', 'PD', 'MR', 'PL', 'GC', 'WM', 'EM'))

# sample tables
all_samples <- d %>%
  group_by(site, collection, type) %>%
  summarize(samples = n() ) %>%
  pivot_wider(names_from = collection, values_from = samples) %>%
    as.data.frame()

h2f_maysep <- d %>%
  filter(species %in% c('G.affinis', 'P.latipinna', 'H.cyanoguttatus', 'L.macrochirus', 'L.megalotis', 'L.gulosus', 'L.cyanellus', 'M.salmoides', 'L.oculatus') |
           type %in% c('Terrestrial', 'Aquatic')) %>%
  filter(collection != '2020-01-15') %>%
  group_by(collection, species, site) %>%
  summarize(samples = n() ) %>%
  mutate(samples = ifelse(samples > 3, 3, samples)) %>%
  ungroup() %>%
  group_by(species, site) %>%
  summarize(samples = sum(samples) ) %>%
  pivot_wider(names_from = site, values_from = samples) %>%
  filter(species != 'clover')

h2f_maysep[is.na(h2f_maysep)] <- 0

h2f_maysep_gtotal <- h2f_maysep %>% 
  rowwise() %>% 
  mutate(myTidySum = sum(c_across(PD:EM)),
         project = 'hydrogen') %>%
  group_by(project) %>%
  summarize(grandtotal = sum(myTidySum))

# The high-end estimate for the number of samples for hydrogen testing would be 347 plus any invertebrate powders that would be recovered and deemed suitable. This grand total includes a maximum of 3 replicates within each of nine species of fish and sources. Species were not distinguished by size. Patterns in mixing models from the 3-site pilot suggest that the basal resource patterns across the precipitation gradient are most evident at higher trophic levels; fish species from P.latipinna to L.gulosus had progressive trends with non-overlapping credible intervals.

# Aggregating seasons: The estimated 347 hydrogen samples includes powders from both summer and autumn sampling periods. A benefit for running both seasons is that not all fish were caught evenly through each sampling period, so an aggregated approach for hydrogen carbon mixing models could provide 9 estimates of % aquatic resources for a given species which could produce regressions. However, aggregating 2 seasons assumes temporal stability in resource isotopes. Given that hydrogen isotopes were more stable in sources spatially, and carbon does not consistently vary across seasons, I think aggregating hydrogen signals across the seasons would provide the most powerful dataset for terrestrial vs aquatic mixing models.

# Invertebrates: If we seek to connect the trophic interactions from upper to intermediate levels, we could run inverterbates. Invertebrate orders that are likely to contain leftover powder would include gastropoda, bivalva, odonata, decapoda, and hemiptera. Also, Hydrogen tins are filled with a much smaller amount of material, so recovering powders although tedious is feasible for more samples than Chris Groff and myself might initially think possible.

# size class: To assess whether basal resource dependency shifts with life-stage of common lepomis taxa, I would recommend running an additional (n = 9 sites, 2 additional Lepomis groups, 3 samples, 2 seasons) 112 samples. Thus, the top-end estimate This would make a total of 559 samples + inverts for hydrogen analysis.

## Available May samples
# missing periphyton AND filamentous algae at WM
# missing gambusia at TR AR MR
# missing gambusia AND poecilids at AR

# Running hydrogen on only May samples would involve about 177 samples + inverts (n~80) + sizeclass fish (n~100)

mafish_available <- d %>%
  filter(species %in% c('G.affinis', 'P.latipinna', 'H.cyanoguttatus', 'L.macrochirus', 'L.megalotis', 'L.gulosus', 'L.cyanellus', 'M.salmoides', 'L.oculatus') ) %>%
  filter(species != 'clover') %>%
  filter(collection == '2020-05-15') %>%
  group_by(species, site) %>%
  summarize(samples = n() ) %>%
  pivot_wider(names_from = site, values_from = samples) %>%
  as.data.frame()
mafish_available[is.na(mafish_available)] <- 0

masrc_available <- d %>%
  filter(type %in% c('Aquatic', 'Terrestrial') ) %>%
  filter(species != 'clover') %>%
  filter(collection == '2020-05-15') %>%
  group_by(type, species, site) %>%
  arrange(type) %>%
  summarize(samples = n() ) %>%
  pivot_wider(names_from = site, values_from = samples) %>%
  as_tibble() %>%
  relocate(TR, .before = SF)
masrc_available[is.na(masrc_available)] <- 0

mafish_total <- mafish_available %>%
  pivot_longer(cols = c(SF:MR), names_to = 'site', values_to = 'n') %>%
  mutate(n_cap = ifelse(n >= 3, 3, n),
         project = 'fish') %>%
  select(-n) %>%
  group_by(project) %>%
  summarize(n_tot = sum(n_cap))

masrc_total <- masrc_available %>%
  pivot_longer(cols = c(TR:EM), names_to = 'site', values_to = 'n') %>%
  mutate(project = 'source') %>%
  group_by(project) %>%
  summarise(n_tot = sum(n))

mafish_available
mafish_total
masrc_available  
masrc_total

## May small-mouthed sunfish (megalotis and macrochirus) size class sample distribution
# missing <50mm at TR MR GC
# missing <100mm at TR
# missing <200mm at SF

# recommend analyzing hydrogen in 100mm (n = 24) and 200mm (25)
May_lepomis_sm <- d %>%
  filter(species %in% c('L.macrochirus', 'L.megalotis') ) %>%
  filter(collection == '2020-05-15') %>%
  mutate(maxfl_bin = ifelse(max_forklength <= 50, '50mm',
                     ifelse(max_forklength > 50 & 
                            max_forklength <= 100 , '100mm',
                     ifelse(max_forklength > 100 &
                            max_forklength <= 200, '200mm',
                     ifelse(max_forklength > 200 &
                            max_forklength <= 300, '300mm',
                     ifelse(max_forklength > 300 &
                            max_forklength <= 500, '500mm', max_forklength)))))) %>%
  group_by(site, maxfl_bin) %>%
  summarize(available = n()) %>%
  pivot_wider(names_from = maxfl_bin, values_from = available) %>%
  relocate('50mm', .before='200mm') %>%
  relocate('100mm', .before='200mm')
May_lepomis_sm[is.na(May_lepomis_sm)] <- 0
May_lepomis_sm 

## May large-mouthed sunfish (megalotis and macrochirus) size class sample distribution
# missing <50mm at TR, SF, MR, PL, GC
# missing <100mm at TR
# missing <200mm at TR, PL, WM

# Recommend analyzing individuals in the <200 (n = 26) and <300 (n = 22)size class range
May_lepomis_lm <- d %>%
  filter(species %in% c('L.gulosus', 'L.cyanellus') ) %>%
  mutate(maxfl_bin = ifelse(max_forklength <= 50, '50mm',
                     ifelse(max_forklength > 50 & 
                            max_forklength <= 100 , '100mm',
                     ifelse(max_forklength > 100 &
                            max_forklength <= 200, '200mm',
                     ifelse(max_forklength > 200 &
                            max_forklength <= 300, '300mm',
                     ifelse(max_forklength > 300 &
                            max_forklength <= 500, '500mm', max_forklength)))))) %>%
  group_by(site, maxfl_bin) %>%
  summarize(available = n()) %>%
  pivot_wider(names_from = maxfl_bin, values_from = available) %>%
  relocate('50mm', .before='100mm')
May_lepomis_lm[is.na(May_lepomis_lm)] <- 0
May_lepomis_lm

## Available September samples
# September sources: missing periphyton AND filamentous algae at PL
# September gambusia missing at TR GC
# Gambusia AND Poecilid missing at GC

# Running only samples from September would entail 128 fish samples + 43 source samples for a total of 173 samples + inverts (3 spp -> n~80) + additional size class samples (n~100)

sefish_available <- d %>%
  filter(species %in% c('G.affinis', 'P.latipinna', 'H.cyanoguttatus', 'L.macrochirus', 'L.megalotis', 'L.gulosus', 'L.cyanellus', 'M.salmoides', 'L.oculatus') ) %>%
  filter(species != 'clover') %>%
  filter(collection == '2020-09-15') %>%
  group_by(species, site) %>%
  summarize(samples = n() ) %>%
  pivot_wider(names_from = site, values_from = samples) %>%
  as.data.frame()
sefish_available[is.na(sefish_available)] <- 0

sesrc_available <- d %>%
  filter(type %in% c('Aquatic', 'Terrestrial') ) %>%
  filter(species != 'clover') %>%
  filter(collection == '2020-09-15') %>%
  group_by(type, species, site) %>%
  arrange(type) %>%
  summarize(samples = n() ) %>%
  pivot_wider(names_from = site, values_from = samples) %>%
  as_tibble() %>%
  relocate(TR, .before = SF)
sesrc_available[is.na(sesrc_available)] <- 0

sefish_total <- sefish_available %>%
  pivot_longer(cols = c(SF:GC), names_to = 'site', values_to = 'n') %>%
  mutate(n_cap = ifelse(n >= 3, 3, n),
         project = 'fish') %>%
  select(-n) %>%
  group_by(project) %>%
  summarize(n_tot = sum(n_cap))

sesrc_total <- sesrc_available %>%
  pivot_longer(cols = c(PD:PL), names_to = 'site', values_to = 'n') %>%
  mutate(project = 'source') %>%
  group_by(project) %>%
  summarise(n_tot = sum(n))

sefish_available
sefish_total
sesrc_available  
sesrc_total


## September small-mouthed sunfish (megalotis and macrochirus) size class sample distribution
#  <50mm missing n at MR GC, low n at PL WM EM
#  <100mm low n at MR
#  <200mm low n at AR PD PL WM EM

# recommend analyzing hydrogen in 100mm (n = 19) and 200mm (13) in small-mouthed Lepomis September
September_lepomis_sm <- d %>%
  filter(species %in% c('L.macrochirus', 'L.megalotis') ) %>%
  filter(collection == '2020-09-15') %>%
  mutate(maxfl_bin = ifelse(max_forklength <= 50, '50mm',
                     ifelse(max_forklength > 50 & 
                            max_forklength <= 100 , '100mm',
                      ifelse(max_forklength > 100 &
                             max_forklength <= 200, '200mm',
                      ifelse(max_forklength > 200 &
                             max_forklength <= 300, '300mm',
                      ifelse(max_forklength > 300 &
                             max_forklength <= 500, '500mm', max_forklength)))))) %>%
  group_by(site, maxfl_bin) %>%
  summarize(available = n()) %>%
  pivot_wider(names_from = maxfl_bin, values_from = available) %>%
  relocate('50mm', .before='200mm') %>%
  relocate('100mm', .before='200mm')
September_lepomis_sm[is.na(September_lepomis_sm)] <- 0
September_lepomis_sm 

## September large-mouthed sunfish (megalotis and macrochirus) size class sample distribution
# <50mm missing n at GC, low n at SF, PD, MR, PL, EM
# <100mm missing n at SF, MR, low n at AR, PL, GC, EM
# <200mm missing n at SF, AR, WM, low n at MR, PL, GC, EM

# Recommend analyzing hydrogen in large-mouthed lepomis in the 100-300 size class range ( n = 18)
September_lepomis_lm <- d %>%
  filter(species %in% c('L.gulosus', 'L.cyanellus') ) %>%
  filter(collection == '2020-09-15') %>%
  mutate(maxfl_bin = ifelse(max_forklength <= 50, '50mm',
                     ifelse(max_forklength > 50 & 
                            max_forklength <= 100 , '100mm',
                      ifelse(max_forklength > 100 &
                             max_forklength <= 200, '200mm',
                      ifelse(max_forklength > 200 &
                             max_forklength <= 300, '300mm',
                      ifelse(max_forklength > 300 &
                             max_forklength <= 500, '500mm', max_forklength)))))) %>%
  group_by(site, maxfl_bin) %>%
  summarize(available = n()) %>%
  pivot_wider(names_from = maxfl_bin, values_from = available) %>%
  relocate('50mm', .before='100mm')
September_lepomis_lm[is.na(September_lepomis_lm)] <- 0
September_lepomis_lm

coarse_vs_leaf <- d %>%
  filter(species %in% c('coarse_detritus', 'tree_leaf') ) %>%
  ggplot(aes(x=site, y = carbon, fill = species)) +
           geom_boxplot()
```

# Problem recap:
The original plan was to analyze carbon and nitrogen and utilize the weight of sample counts to overcome the relatively narrow difference in carbon values for these systems. However, the outcome of the study was few counts of samples from terrestrial and aquatic sources. By analyzing what few sources we have, we could greatly improve the statistical resolution of our mixing models. This ultimately improves our ability to accurately find the threshold along the precipitation gradient where basal resource acquisition in aquatic food webs change.

# Available samples from May + September
```{r}
h2f_maysep
h2f_maysep_gtotal 
```
### Top end estimate: <610 samples

The high-end estimate for the number of samples for hydrogen testing would be 347 plus any invertebrate powders (n~100) that would be recovered and deemed suitable AND additional replicates within size-classes for 2 lepomis groups (n~160). 347 includes a maximum of 3 replicates within each of nine species of fish and sources. Species were not distinguished by size. Patterns in mixing models from the 3-site pilot suggest that the basal resource patterns across the precipitation gradient are most evident at higher trophic levels; fish species from P.latipinna to L.gulosus had progressive trends with non-overlapping credible intervals. So invertebrate samples could be left out of the hydrogen analysis if our questions remain broad.

### Aggregating seasons: 

The estimated 347 hydrogen samples includes powders from both summer and autumn sampling periods. A benefit for running both seasons is that not all fish were caught evenly through each sampling period, so an aggregated approach for hydrogen carbon mixing models could provide 9 estimates of % aquatic resources for a given species which could produce regressions. However, aggregating 2 seasons assumes temporal stability in resource isotopes. Given that hydrogen isotopes were more stable in sources spatially, and carbon does not consistently vary across seasons, I think aggregating hydrogen signals across the seasons would provide the most powerful dataset for terrestrial vs aquatic mixing models.

### Invertebrates:

If we seek to connect the trophic interactions from upper to intermediate levels, we could run inverterbates. Invertebrate orders that are likely to contain leftover powder would include gastropoda, bivalva, odonata, decapoda, and hemiptera. Also, Hydrogen tins are filled with a much smaller amount of material, so recovering powders although tedious is feasible for more samples than Chris Groff and myself might initially think possible.

### size class: 

To assess whether basal resource dependency shifts with life-stage of common lepomis taxa, I would recommend running an additional (n = 9 sites, 2 additional Lepomis groups, 3 samples, 2 seasons) 112 samples. Thus, the top-end estimate This would make a total of 559 samples + inverts for hydrogen analysis.

# Available May samples

Running hydrogen on only May samples would involve about 177 samples + inverts (n~80) + sizeclass fish (n~100)

### May sources
missing periphyton AND filamentous algae at WM
```{r}
masrc_available  
masrc_total
```

### May fish
missing gambusia at TR AR MR
missing gambusia AND poecilids at AR
```{r}
mafish_available
mafish_total
```


### May small-mouthed sunfish (megalotis and macrochirus) size class sample distribution
```{r}
May_lepomis_sm 
```
missing <50mm at TR MR GC
missing <100mm at TR
missing <200mm at SF



### May large-mouthed sunfish (megalotis and macrochirus) size class sample distribution
```{r}
May_lepomis_lm
```
missing <50mm at TR, SF, MR, PL, GC
missing <100mm at TR
missing <200mm at TR, PL, WM


# Available September samples

September gambusia missing at TR GC
Gambusia AND Poecilid missing at GC

Running only samples from September would entail 128 fish samples + 43 source samples for a total of 173 samples + inverts (3 spp -> n~80) + additional size class samples (n~100)

# The problem with May samples

### May 'Coarse Detritus' aka CPOM'
In the May samples, few leaves were collected, and coarse detritus was considered to be a terrestrial source. While that is reasonable to assume since it the material is originally derived from terrestrial sources, CPOM is contaminated with aquatic by nature of aquatic decomposition and colonization by periphyton. The following figure depicts how the carbon signature of coarse detritus is between the periphyton and tree leaves.

```{r}
coarse_vs_leaf
```




## September fish
```{r}
sefish_available
sefish_total
```

## September Sources
September sources: missing periphyton AND filamentous algae at PL
```{r}
sesrc_available  
sesrc_total
```


### September small-mouthed sunfish (megalotis and macrochirus) size class sample distribution
```{r}
September_lepomis_sm 
```
<50mm missing n at MR GC, low n at PL WM EM
<100mm low n at MR
<200mm low n at AR PD PL WM EM



### September large-mouthed sunfish (megalotis and macrochirus) size class sample distribution
```{r}
September_lepomis_lm
```
<50mm missing n at GC, low n at SF, PD, MR, PL, EM
<100mm missing n at SF, MR, low n at AR, PL, GC, EM
<200mm missing n at SF, AR, WM, low n at MR, PL, GC, EM



# Conclusion:

## September only
If we merely seek the threshold for basal resource shift, I recommend the conservative hydrogen analysis of September samples

.A conservative hydrogen analysis should use the samples from September because we May has few terrestrial sources to sample. This would entail 128 fish samples + 43 source samples for a total of 173 samples + inverts (3 spp -> n~80) + additional size class samples (n~100).


## Sep + May
If we want to look deeper within the food web linkages, I'd recommend using samples from both May and September with options of invertebrate and size-classes.

A more ambitious hydrogen analysis should include both May and September samples which would entail <610 samples: 347 plus any invertebrate powders (n~100) that would be recovered and deemed suitable AND additional replicates within size-classes for 2 lepomis groups (n~160).











